### 2014-11-05 implementation.tex
edits

Replicated data types and the stronger consistency semantics are implemented
and enforced in the \emph{shim layer}. Our implementation supports eventual,
causal, and strong consistency for data type operations, and RC, MAV, and RR
semantics for transactions.  This functionality is implemented entirely on top
of the standard interface exposed by Cassandra. From an engineering
perspective, leveraging an off-the-shelf data store enables an implementation
comprising roughly only 2500 lines of Haskell code, packaged as a
library\footnote{url to the hackage}.

%% Complex business logic is implemented with the help of
%% transactions on the web-server side, with clients interacting with the
%% web-service over REST interface.

@@

\noindent Since the semantics of summarization depends on the semantics of the data
type, our programming model provides a summarization function for each RDT:
\begin{codehaskell}
summarize :: Effect e => [e] -> [e]
\end{codehaskell}
\noindent with the intention that the length of the result is smaller that the
length of the argument. We utilize the \cf{summarize} function to summarize the
object state both in the shim layer node and the backing store, typically when
the number of effects on an object crosses a tunable threshold. Shim layer
summarization is straight-forward; a summarization thread takes the local lock
on the object, and replaces its state with the summarized state. The shim layer
node remains unavailable for that particular object during summarization
(usually a few milliseconds).

>>>>

Replicated data types and the stronger consistency semantics are implemented
and enforced in the \emph{shim layer}. Our implementation supports eventual,
causal, and strong consistency for data type operations, and RC, MAV, and RR
semantics for transactions.  This functionality is implemented entirely on top
of the standard interface exposed by Cassandra. From an engineering
perspective, leveraging an off-the-shelf data store enables an implementation
comprising roughly only 2500 lines of Haskell code, which is packaged as a
library.

@@

Since the semantics of summarization depends on the semantics of the data type,
our programming model provides a summarization function (\cf{summarize}) for
each RDT of type \cf{[e] -> [e]} with the intention that the length of the
result is smaller that the length of the argument. We utilize the
\cf{summarize} function to summarize the object state both in the shim layer
node and the backing store, typically when the number of effects on an object
crosses a tunable threshold. Shim layer summarization is straight-forward; a
summarization thread takes the local lock on the object, and replaces its state
with the summarized state. The shim layer node remains unavailable for that
particular object during summarization (usually a few milliseconds).

### 2014-11-05 transactions.tex
edits

While contracts on individual operations offer the programmer object-level
declarative reasoning, real-world scenarios often involve operations that span
multiple objects. In order to address this problem, several recent
systems~\cite{Walter,Burckhardt2012,BailisHAT} have proposed a variety of
transactions, with varying semantics, in order to compose operations on
multiple objects. However, given that classical transaction models such as
serializability and snapshot isolation require inter-replica coordination,
these systems espouse \emph{coordination-free transactions} that remain
available under network partitions, but only provide weaker isolation
guarantees. Coordination-free transactions have intricate consistency semantics
and widely varying runtime overheads. As with operation-level consistency, the
onus is on the programmer to pick the correct transaction kind; this selection
is complicated by operation consistency requirements.

@@

We assume that operations not part of any transaction belong to their own
unique transaction. While transactions may have varying isolation
guarantees, we make the standard assumption that all transactions provide
atomicity. Hence, we include the following axiom in $\Delta$: \[\small \forall
a,b,c.~\txn{a}{b,c} ~\wedge~ \sameobj{b}{c} ~\wedge~ \vis{b}{a} \Rightarrow
\vis{c}{a}\] \noindent The semantics of this contract is illustrated in
Figure~\ref{fig:txn_atomicity}.

@@

\noindent \begin{minipage}[t]{0.53\columnwidth}
\begin{codehaskell}
save amt =
  x <- dollar(classify $\psi_1$)
  atomically x dollar do
    b <- withdraw c amt
    when b dollar deposit s amt
\end{codehaskell}
\end{minipage}
\begin{minipage}[t]{0.47\columnwidth}
\begin{codehaskell}
totalBalance =
  x <- dollar(classify $\psi_2$)
  atomically x dollar do
    b1 <- getBalance s
    b2 <- getBalance s
    return b1 + b2
\end{codehaskell}
\end{minipage}

\noindent $\cv_1$ and $\cv_2$ are the contracts on the corresponding
transactions. The function \cf{classify} assigns the contracts
\emph{statically} to one of the transaction isolation levels offered by the
store; \cf{\$()} is meta-programming syntax for splicing the result into the
program. The \cf{atomically} construct invokes the enclosing operations at the
given isolation level $x$, ensuring that the effects of the operations are made
visible atomically. Our goal is to ensure that \cf{totalBalance} returns the
result obtained from a consistent snapshot of the object states.

While making both transactions serializable would ensure correctness,
distributed stores rarely offer serializable transactions. Moreover,
serializability is unavailable and hinders scalability. As we will see, these
transactions can be satisfied with much weaker isolation guarantees. Despite
the atomicity offered by the transaction, anomalies are still possible. For
example, the two \cf{getBalance} operations in \cf{totalBalance} transactions
might be served by different replicas with distinct set of committed \cf{save}
transactions. If the first(second) \cf{getBalance} operation witness a
\cf{save} transaction that is not witnessed by the second(first)
\cf{getBalance} operation, then the balance returned will be less(greater) than
the actual balance. It is not immediately apparent which weakest isolation
guarantee will be sufficient to prevent the anomaly.

Instead, \name requires the programmer to simply state the consistency
requirement as a contract. Since we would like both the \cf{getBalance}
operations to witness the same set of \cf{save} transactions, we define the
constraint on \cf{totalBalance} transaction $\psi_2$ as:

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\cv_{2} = \forall a:\rcf{getBalance}, b:\rcf{getBalance}, \\
\qquad (c:\rcf{withdraw} \vee \rcf{deposit}), (c:\rcf{withdraw} \vee \rcf{deposit}). \\
\qquad \txn{a,b}{c,d} ~\wedge~ \vis{c}{a} ~\wedge~ \sameobj{d}{b} \Rightarrow \vis{d}{b}
\end{array}
\end{smathpar}

\noindent The key idea in the above definition is that the $\txnZ$ primitive
allows us to relate operations on different objects.

The \cf{save} transaction only needs to ensure that the two writes it performs
are made visible atomically. Since this is ensured by combining them in a
transaction, \cf{save} does not require any additional constraints, and
$\psi_1$ is simply $\small \true$.

@@

A transaction with ANSI RC semantics only witnesses committed operations. Let
us assume that the store buffers updates until all the updates from the
transaction are available at a replica. If the transaction commits, the
buffered updates are made visible. Otherwise, the buffered updates are
discarded. In particular, a store implementing RC does not require
inter-replica coordination. RC does not entail any further isolation
guarantees. We can express RC as follows:

\vspace{-1em}
\begin{smathpar}
\rcc = \forall a,b,c.~\txn{a}{b,c} \wedge \sameobj{b}{c} \wedge \vis{b}{a} \Rightarrow \vis{c}{a}
\end{smathpar}

@@

RR semantics requires that the transaction witness a snapshot of the data store
state. Importantly, this snapshot can be obtained from any replica, and hence
RR is coordination-free. An example for such a transaction is the
\cf{totalBalance} transaction. The semantics of RR is captured by the following
contract:

>>>>

While contracts on individual operations offer the programmer object-level
declarative reasoning, real-world scenarios often involve operations that span
multiple objects. In order to address this problem, several recent
systems~\cite{Walter,Burckhardt2012,BailisHAT} have proposed eventually
consistent transactions in order to compose operations on multiple objects.
However, given that classical transaction models such as serializability and
snapshot isolation require inter-replica coordination, these systems espouse
\emph{coordination-free transactions} that remain available under network
partitions, but only provide weaker isolation guarantees. Coordination-free
transactions have intricate consistency semantics and widely varying runtime
overheads. As with operation-level consistency, the onus is on the programmer
to pick the correct transaction kind. This choice is further is complicated by
consistency semantics of individual operations.

@@

We assume that operations not part of any transaction belong to their own
unique transaction. While transactions may have varying isolation guarantees,
we make the standard assumption that all transactions provide atomicity. Hence,
we include the following axiom in $\Delta$:

\vspace{-1.5em}
\begin{smathpar}
\forall a,b,c.~\txn{a}{b,c} ~\wedge~ \sameobj{b}{c} ~\wedge~ \vis{b}{a}
\Rightarrow \vis{c}{a}
\end{smathpar}

\noindent The semantics of this contract is illustrated in
Figure~\ref{fig:txn_atomicity}.

@@

\noindent \begin{minipage}[t]{0.53\columnwidth}
\begin{codehaskell}
save amt =
  x <- dollar(classify $\cv_{sv}$)
  atomically x dollar do
    b <- withdraw c amt
    when b dollar deposit s amt
\end{codehaskell}
\end{minipage}
\begin{minipage}[t]{0.47\columnwidth}
\begin{codehaskell}
totalBalance =
  x <- dollar(classify $\cv_{tb}$)
  atomically x dollar do
    b1 <- getBalance s
    b2 <- getBalance s
    return b1 + b2
\end{codehaskell}
\end{minipage}

\noindent $\cv_{sv}$ and $\cv_{tb}$ are the contracts on the corresponding
transactions. The function \cf{classify} assigns the contracts
\emph{statically} to one of the transaction isolation levels offered by the
store; \cf{\$()} is meta-programming syntax for splicing the result into the
program. The \cf{atomically} construct invokes the enclosing operations at the
given isolation level $x$, ensuring that the effects of the operations are made
visible atomically. Our goal is to ensure that \cf{totalBalance} returns the
result obtained from a consistent snapshot of the object states.

While making both transactions serializable would ensure correctness,
distributed stores rarely offer serializable transactions since it is
unavailable and hinders scalability~\cite{BailisHAT}. As we will see, these
transactions can be satisfied with much weaker isolation guarantees. Despite
the atomicity offered by the transaction, anomalies are still possible. For
example, the two \cf{getBalance} operations in \cf{totalBalance} transactions
might be served by different replicas with distinct set of committed \cf{save}
transactions. If the first(second) \cf{getBalance} operation witness a
\cf{save} transaction that is not witnessed by the second(first)
\cf{getBalance} operation, then the balance returned will be less(greater) than
the actual balance. It is not immediately apparent which weakest isolation
guarantee will be sufficient to prevent the anomaly.

Instead, \name requires the programmer to simply state the consistency
requirement as a contract. Since we would like both the \cf{getBalance}
operations to witness the same set of \cf{save} transactions, we define the
constraint on \cf{totalBalance} transaction $\psi_{tb}$ as:

\vspace{-1.2em}
\begin{smathpar}
\begin{array}{l}
\cv_{tb} = \forall a:\rcf{getBalance}, b:\rcf{getBalance}, \\
\qquad (c:\rcf{withdraw} \vee \rcf{deposit}), (d:\rcf{withdraw} \vee \rcf{deposit}). \\
\qquad \txn{a,b}{c,d} ~\wedge~ \vis{c}{a} ~\wedge~ \sameobj{d}{b} \Rightarrow \vis{d}{b}
\end{array}
\end{smathpar}

\noindent The key idea in the above definition is that the $\txnZ$ primitive
allows us to relate operations on different objects. The \cf{save} transaction
only needs to ensure that the two writes it performs are made visible
atomically. Since this is ensured by combining them in a transaction, \cf{save}
does not require any additional constraints, and $\psi_1$ is simply $\small
\true$.

@@

A transaction with ANSI RC semantics only witnesses committed operations. Let
us assume that the store buffers updates until all the updates from the
transaction are available at a replica. If the transaction commits, the
buffered updates are made visible. Otherwise, the buffered updates are
discarded. RC does not entail any further isolation guarantees. Hence, a store
implementing RC does not require inter-replica coordination. We can express RC
as follows:

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\rcc = \forall a,b,c.~\txn{a}{b,c} \wedge \sameobj{b}{c} \\
\qquad \qquad \qquad \wedge~ \vis{b}{a} \Rightarrow \vis{c}{a}
\end{array}
\end{smathpar}

@@

ANSI RR semantics requires that the transaction witness a snapshot of the data
store state. Importantly, this snapshot can be obtained from any replica, and
hence RR is coordination-free. An example for such a transaction is the
\cf{totalBalance} transaction. The semantics of RR is captured by the following
contract:
